---
title: "GSE111672 PDAC-A CopyKAT and UBE2H analysis"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, message=FALSE, warning=FALSE}
library(data.table)
library(Matrix)
library(tidyverse)
library(copykat)
library(Seurat)
library(ComplexHeatmap)
library(circlize)
library(grid)
library(ggpubr)
library(gt)
library(broom)
library(svglite)
library(here)

dir.create(here("figures"), recursive = TRUE, showWarnings = FALSE)
patient <- "PDAC-A"
counts_file <- here("data", "GSE111672", "GSE111672_PDAC-A-indrop-filtered-expMat.txt")
output_prefix <- "GSE111672_PDAC_A_inDrop_copykat"
copykat_rds <- here("data", "GSE111672", "GSE111672_PDAC_A_inDrop_copykat.rds")

exclude_arms <- c("13p", "14p", "15p", "21p", "22p")
```

# Read the patient-specific inDrop matrix

```{r read-expmat}
if (!file.exists(counts_file)) {
  stop("Counts file not found: ", counts_file)
}

message("Reading ", patient, " expMat ...")
df <- data.table::fread(
  counts_file,
  sep = "\t",
  header = TRUE,
  data.table = FALSE,
  check.names = FALSE
)

cell_type <- trimws(gsub("\r", "", colnames(df)[-1]))
cell_ids <- paste0("PDAC_A_cell_", seq_along(cell_type))

genes <- trimws(gsub("\r", "", as.character(df[[1]])))
keep_gene <- !is.na(genes) & nzchar(genes) & !duplicated(genes)

mat <- as.matrix(df[keep_gene, -1, drop = FALSE])
storage.mode(mat) <- "integer"
rownames(mat) <- genes[keep_gene]
colnames(mat) <- cell_ids
rm(df)

drop_rbc <- grepl("^RBC", cell_type, ignore.case = TRUE)
mat <- mat[, !drop_rbc, drop = FALSE]
cell_type <- cell_type[!drop_rbc]
cell_ids <- cell_ids[!drop_rbc]

stopifnot(identical(colnames(mat), cell_ids))

cell_metadata <- data.frame(
  cell = cell_ids,
  cell_type = cell_type,
  patient = patient,
  row.names = cell_ids,
  stringsAsFactors = FALSE,
  check.names = FALSE
)

cat(sprintf(
  "%s clean matrix: %d genes x %d cells (RBCs removed)\n",
  patient, nrow(mat), ncol(mat)
))
print(sort(table(cell_metadata$cell_type), decreasing = TRUE))
```

# Define patient-specific reference and observation groups

```{r define-groups}
all_labels <- sort(unique(cell_metadata$cell_type))
obs_pattern <- "Ductal|Cancer"

ref_group_names <- all_labels[!grepl(obs_pattern, all_labels, ignore.case = TRUE)]
obs_group_names <- all_labels[grepl(obs_pattern, all_labels, ignore.case = TRUE)]

norm_cells <- rownames(cell_metadata)[cell_metadata$cell_type %in% ref_group_names]
obs_cells <- rownames(cell_metadata)[cell_metadata$cell_type %in% obs_group_names]

if (length(norm_cells) < 5) {
  stop("Fewer than five reference cells were identified for ", patient)
}
if (length(obs_cells) == 0) {
  stop("No Ductal/Cancer observation cells were identified for ", patient)
}

cat("Reference groups for ", patient, ":\n", sep = "")
print(sort(table(cell_metadata[norm_cells, "cell_type"]), decreasing = TRUE))

cat("\nObservation groups for ", patient, ":\n", sep = "")
print(sort(table(cell_metadata[obs_cells, "cell_type"]), decreasing = TRUE))
```

# Load CopyKAT

```{r copykat}
res <- readRDS(copykat_rds)
names(res)
str(res, max.level = 1)
```

# Check CopyKAT predictions

```{r prediction-qc}
pred_all <- as.data.frame(res$prediction) %>%
  mutate(
    cell_type = cell_metadata[cell.names, "cell_type"],
    input_group = ifelse(cell.names %in% norm_cells, "reference", "observation"),
    patient = patient
  )

cat("Overall calls:\n")
print(table(pred_all$copykat.pred, useNA = "ifany"))

cat("\nCalls by input group:\n")
print(table(pred_all$input_group, pred_all$copykat.pred, useNA = "ifany"))

cat("\nCalls by cell type:\n")
copykat_table <- table(pred_all$cell_type, pred_all$copykat.pred, useNA = "ifany") |>
  as.data.frame.matrix() |>
  tibble::rownames_to_column("Cell type")

copykat_table |>
  gt() |>
  tab_header(title = "CopyKAT predictions by cell type") |>
  cols_label(`Cell type` = "Cell type")
```

# Extract CopyKAT CN values

```{r cna-matrix}
cn <- as.data.frame(res$CNAmat)

required_annotation_cols <- c("chrom", "chrompos", "abspos")
stopifnot(all(required_annotation_cols %in% colnames(cn)))

copykat_cell_cols <- base::intersect(pred_all$cell.names, colnames(cn))
stopifnot(length(copykat_cell_cols) > 0)

cn_values <- as.matrix(cn[, copykat_cell_cols, drop = FALSE])
storage.mode(cn_values) <- "numeric"

cat("CN matrix:", nrow(cn_values), "bins x", ncol(cn_values), "cells\n")
print(range(cn_values, na.rm = TRUE))
print(quantile(cn_values, probs = c(0, 0.01, 0.05, 0.5, 0.95, 0.99, 1), na.rm = TRUE))
```

# Derive the bin-signal threshold from confidently-diploid cells

```{r diploid-threshold}
diploid_cells <- pred_all$cell.names[pred_all$copykat.pred == "diploid"]
diploid_cells <- intersect(diploid_cells, colnames(cn_values))

diploid_signal <- as.numeric(as.matrix(cn_values[, diploid_cells, drop = FALSE]))
diploid_signal <- diploid_signal[is.finite(diploid_signal)]

quantile(abs(diploid_signal), probs = c(0.75, 0.80, 0.90, 0.95, 0.975), na.rm = TRUE)

bin_signal_threshold <- unname(quantile(abs(diploid_signal), probs = 0.90, na.rm = TRUE))
bin_signal_threshold
```

```{r threshold-distribution}
signal_df <- data.frame(signal = diploid_signal)

ggplot(signal_df, aes(x = signal)) +
  geom_histogram(bins = 150, boundary = 0) +
  geom_vline(xintercept = c(-bin_signal_threshold, bin_signal_threshold),
             linetype = "dashed", linewidth = 0.8) +
  labs(
    title = "Distribution of CopyKAT bin signals",
    subtitle = paste0("90th percentile of absolute signal = \u00b1", round(bin_signal_threshold, 3)),
    x = "CopyKAT bin signal", y = "Bin-cell observations"
  ) +
  theme_classic()
```

# Bin-level heatmap by cell type

```{r bin-heatmap-prep}
cn_df <- res$CNAmat
stopifnot(all(c("chrom", "chrompos", "abspos") %in% colnames(cn_df)))

cell_cols <- intersect(colnames(cn_df), res$prediction$cell.names)
stopifnot(length(cell_cols) > 0)

cn_mat_plot <- as.matrix(cn_df[, cell_cols, drop = FALSE])
storage.mode(cn_mat_plot) <- "numeric"

bin_chr <- as.character(cn_df$chrom)
stopifnot(nrow(cn_mat_plot) == length(bin_chr),
          all(colnames(cn_mat_plot) %in% rownames(cell_metadata)))

cell_meta <- data.frame(
  cell = colnames(cn_mat_plot),
  cell_type = as.character(cell_metadata[colnames(cn_mat_plot), "cell_type", drop = TRUE]),
  stringsAsFactors = FALSE
)
rownames(cell_meta) <- cell_meta$cell

stopifnot(identical(rownames(cell_meta), colnames(cn_mat_plot)), !anyNA(cell_meta$cell_type))

cell_type_order <- unique(cell_meta$cell_type)
cell_meta$cell_type <- factor(cell_meta$cell_type, levels = cell_type_order)

cell_order <- cell_meta %>% arrange(cell_type) %>% pull(cell)
cn_mat_plot <- cn_mat_plot[, cell_order, drop = FALSE]
cell_meta <- cell_meta[cell_order, , drop = FALSE]

stopifnot(identical(colnames(cn_mat_plot), rownames(cell_meta)))
```

```{r bin-heatmap-draw, fig.width=12, fig.height=8}
# Validation plot
cell_type_colors <- setNames(
  grDevices::hcl.colors(n = nlevels(cell_meta$cell_type), palette = "Dark 3"),
  levels(cell_meta$cell_type)
)

top_anno <- HeatmapAnnotation(
  `Cell type` = cell_meta$cell_type,
  col = list(`Cell type` = cell_type_colors),
  annotation_name_side = "left",
  show_annotation_name = TRUE,
  annotation_legend_param = list(
    `Cell type` = list(title = "Cell type", labels_gp = gpar(fontsize = 8),
                        title_gp = gpar(fontsize = 9, fontface = "bold"))
  )
)

copykat_col_fun <- circlize::colorRamp2(c(-0.5, 0, 0.5), c("#3B6FB6", "white", "#C74440"))

keep_bins <- bin_chr != "23"
cn_mat_plot <- cn_mat_plot[keep_bins, , drop = FALSE]
bin_chr <- bin_chr[keep_bins]
bin_chr_factor <- factor(bin_chr, levels = as.character(1:22))

ht <- Heatmap(
  cn_mat_plot,
  name = "CopyKAT",
  col = copykat_col_fun,
  top_annotation = top_anno,
  cluster_rows = FALSE,
  row_split = bin_chr_factor,
  cluster_row_slices = FALSE,
  row_gap = unit(0.4, "mm"),
  cluster_columns = FALSE,
  show_row_names = FALSE,
  show_column_names = FALSE,
  row_title = as.character(1:22),
  row_title_rot = 0,
  row_title_gp = grid::gpar(fontsize = 9),
  column_split = cell_meta$cell_type,
  column_gap = unit(2, "mm"),
  rect_gp = gpar(col = NA),
  heatmap_legend_param = list(title = "CopyKAT", at = c(-0.5, 0, 0.5)),
  column_title = "Cells",
  use_raster = FALSE
)

png(filename = here("figures", "GSE111672_PDAC_A_copykat_bin_heatmap_by_cell_type.png"),
    width = 4200, height = 2700, res = 300)
ComplexHeatmap::draw(ht, heatmap_legend_side = "right", annotation_legend_side = "right")
dev.off()

ht
```

# Normalize UBE2H expression

```{r normalize-ube2h}
seu <- CreateSeuratObject(
  counts = as(mat, "dgCMatrix"),
  meta.data = cell_metadata,
  project = paste0("GSE111672_", "PDAC_A"),
  min.cells = 0,
  min.features = 0
)

seu <- NormalizeData(seu, normalization.method = "LogNormalize", scale.factor = 10000, verbose = FALSE)

ube2h_matches <- which(toupper(rownames(seu)) == "UBE2H")
stopifnot(length(ube2h_matches) == 1)
ube2h_id <- rownames(seu)[ube2h_matches]

ube2h_df <- FetchData(seu, vars = ube2h_id, layer = "data")
colnames(ube2h_df) <- "UBE2H_normalized"
ube2h_df$cell <- rownames(ube2h_df)
```

# Build chromosome-arm coordinates

```{r arm-coordinates}
arm_coords <- read_csv(here("data", "arm_coords.csv"))
```

# Assign CopyKAT bins to chromosome arms

```{r assign-bins-to-arms}
bin_annot <- cn %>%
  transmute(
    bin_id = seq_len(nrow(cn)),
    chrom = as.numeric(chrom),
    chrompos = as.numeric(chrompos),
    abspos = as.numeric(abspos)
  )

bin_arm <- bin_annot %>%
  inner_join(
    arm_coords %>% transmute(chrom_num, arm, arm_start = start, arm_end = end),
    by = join_by(chrom == chrom_num, chrompos >= arm_start, chrompos < arm_end)
  ) %>%
  filter(!arm %in% exclude_arms) %>%
  dplyr::select(bin_id, chrom, chrompos, abspos, arm)

assignment_qc <- bin_arm %>%
  count(bin_id) %>%
  summarise(max_assignments = max(n), duplicated_bins = sum(n > 1))
print(assignment_qc)
stopifnot(assignment_qc$max_assignments <= 1)
```

# Arm-level calls and aneuploidy score

An arm is called gained or lost when at least 80% of its CopyKAT genomic bins cross the selected signal threshold (derived above from confidently-diploid cells). 
The short acrocentric p arms are excluded before arm calling and scoring.

```{r arm-calls}
print(bin_signal_threshold)
fraction_threshold <- 0.80

arms <- unique(bin_arm$arm)

arm_call <- matrix(0L, nrow = length(arms), ncol = ncol(cn_values),
                    dimnames = list(arms, colnames(cn_values)))

for (a in arms) {
  bin_ids <- bin_arm$bin_id[bin_arm$arm == a]
  x <- cn_values[bin_ids, , drop = FALSE]

  frac_gain <- colMeans(x > bin_signal_threshold, na.rm = TRUE)
  frac_loss <- colMeans(x < -bin_signal_threshold, na.rm = TRUE)

  arm_call[a, frac_gain >= fraction_threshold] <- 1L
  arm_call[a, frac_loss >= fraction_threshold] <- -1L
}

stopifnot(!any(rownames(arm_call) %in% exclude_arms))
stopifnot(all(arm_call %in% c(-1L, 0L, 1L)))

aneuploidy_score <- colSums(abs(arm_call))
summary(aneuploidy_score)
```

# Arm-call heatmap, by cell type (validates the custom scoring method)

```{r arm-call-heatmap, fig.width=12, fig.height=8}
arm_order <- as.vector(rbind(paste0(1:22, "p"), paste0(1:22, "q")))
arm_order <- arm_order[arm_order %in% rownames(arm_call)]
arm_call_plot <- arm_call[arm_order, , drop = FALSE]

heatmap_meta <- pred_all %>%
  transmute(cell = cell.names, cell_type, copykat_call = copykat.pred, input_group) %>%
  filter(cell %in% colnames(arm_call_plot))

rownames(heatmap_meta) <- heatmap_meta$cell
heatmap_meta <- heatmap_meta[colnames(arm_call_plot), , drop = FALSE]

stopifnot(identical(rownames(heatmap_meta), colnames(arm_call_plot)))

heatmap_meta$aneuploidy_score <- colSums(abs(arm_call_plot))

cell_type_levels <- unique(as.character(heatmap_meta$cell_type))
heatmap_meta <- heatmap_meta %>%
  mutate(cell_type = factor(as.character(cell_type), levels = cell_type_levels))

cell_order <- heatmap_meta %>% arrange(cell_type, aneuploidy_score) %>% pull(cell)
arm_call_plot <- arm_call_plot[, cell_order, drop = FALSE]
heatmap_meta <- heatmap_meta[cell_order, , drop = FALSE]

stopifnot(identical(colnames(arm_call_plot), rownames(heatmap_meta)))

call_colors <- c(`-1` = "#3B6FB6", `0` = "white", `1` = "#C74440")
cell_type_colors_arm <- setNames(
  grDevices::hcl.colors(n = length(cell_type_levels), palette = "Dark 3"),
  cell_type_levels
)

top_anno_arm <- HeatmapAnnotation(
  `Cell type` = heatmap_meta$cell_type,
  `Arm score` = anno_barplot(heatmap_meta$aneuploidy_score, border = FALSE, height = unit(1.5, "cm")),
  col = list(`Cell type` = cell_type_colors_arm),
  annotation_name_side = "left",
  annotation_legend_param = list(
    `Cell type` = list(title = "Cell type", labels_gp = gpar(fontsize = 8),
                        title_gp = gpar(fontsize = 9, fontface = "bold"))
  )
)

ht_arm <- Heatmap(
  arm_call_plot,
  name = "Arm call",
  col = call_colors,
  top_annotation = top_anno_arm,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_row_names = TRUE,
  show_column_names = FALSE,
  row_names_side = "left",
  row_names_gp = gpar(fontsize = 9),
  column_split = heatmap_meta$cell_type,
  column_gap = unit(2, "mm"),
  column_title = NULL,
  rect_gp = gpar(col = NA),
  heatmap_legend_param = list(title = "Arm call", at = c(-1, 0, 1),
                               labels = c("Loss", "Neutral", "Gain")),
  row_title = "Chromosome arm"
)

png(filename = here("figures", paste0(output_prefix, "_arm_heatmap_by_cell_type.png")),
    width = 4200, height = 2700, res = 300)
ComplexHeatmap::draw(ht_arm, heatmap_legend_side = "right", annotation_legend_side = "right")
dev.off()

ht_arm
```

# Assemble cell-level metadata for downstream UBE2H and aneuploidy analyses

```{r assemble-final-metadata}
final_meta <- pred_all %>%
  transmute(cell = cell.names, cell_type, copykat_call = copykat.pred, input_group) %>%
  left_join(ube2h_df, by = "cell") %>%
  mutate(aneuploidy_score = .env$aneuploidy_score[match(cell, names(.env$aneuploidy_score))]) %>%
  filter(input_group == "observation", !is.na(aneuploidy_score), !is.na(UBE2H_normalized))
```

# Chromosome 7q status

```{r chr7q-status}
stopifnot("7q" %in% rownames(arm_call))

final_meta <- final_meta %>%
  mutate(
    chr7q_call = arm_call["7q", match(cell, colnames(arm_call))],
    chr7q_status = factor(chr7q_call, levels = c(-1, 0, 1), labels = c("Loss", "Neutral", "Gain"))
  )

plot_df_7q <- final_meta %>%
  filter(input_group == "observation", !is.na(UBE2H_normalized), !is.na(chr7q_status))

print(table(plot_df_7q$chr7q_status, useNA = "ifany"))

if (nlevels(droplevels(plot_df_7q$chr7q_status)) >= 2) {
  pairwise_results_7q <- compare_means(
    UBE2H_normalized ~ chr7q_status, data = plot_df_7q,
    method = "wilcox.test", p.adjust.method = "BH"
  )
  pairwise_results_7q %>% gt()
}
```

# Aneuploidy score excluding chromosome 7q

This removes 7q from the burden, preventing the chromosome arm carrying UBE2H from contributing to the predictor.

```{r score-without-chr7q}
chr7q_arm <- intersect("7q", rownames(arm_call))
arm_call_no_chr7q <- arm_call[!rownames(arm_call) %in% chr7q_arm, , drop = FALSE]

aneuploidy_score_no_chr7q <- colSums(abs(arm_call_no_chr7q))

lm_df <- final_meta %>%
  mutate(
    aneuploidy_score_no_chr7q = .env$aneuploidy_score_no_chr7q[match(cell, names(.env$aneuploidy_score_no_chr7q))],
    chr7q_status = relevel(factor(chr7q_status), ref = "Neutral")
  ) %>%
  filter(input_group == "observation", !is.na(UBE2H_normalized),
         !is.na(aneuploidy_score_no_chr7q), !is.na(chr7q_status))

cor_no_chr7q <- cor.test(lm_df$aneuploidy_score_no_chr7q, lm_df$UBE2H_normalized,
                         method = "spearman", exact = FALSE)
cor_no_chr7q

fit_7q_adjusted <- lm(UBE2H_normalized ~ aneuploidy_score_no_chr7q + chr7q_status, data = lm_df)
summary(fit_7q_adjusted)
broom::tidy(fit_7q_adjusted, conf.int = TRUE) %>% gt()
```

# UBE2H by fixed aneuploidy-score groups (cancer cells, chr7q-neutral only)

```{r fixed-score-bin-plot, fig.width=7, fig.height=5.5}
# Cancer cells only, chromosome 7q-neutral only
pdac_plot_df <- lm_df %>%
  filter(
    cell_type %in% c("Cancer clone A", "Cancer clone B"),
    chr7q_status == "Neutral",
    !is.na(aneuploidy_score_no_chr7q),
    !is.na(UBE2H_normalized)
  )

cat(
  "Final plotting subset:",
  nrow(pdac_plot_df),
  "cells\n"
)

cat(
  "Aneuploidy-score range:",
  min(pdac_plot_df$aneuploidy_score_no_chr7q, na.rm = TRUE),
  "to",
  max(pdac_plot_df$aneuploidy_score_no_chr7q, na.rm = TRUE),
  "\n"
)

# Spearman correlation is calculated using the original continuous, cell-level aneuploidy score. 
# Score groups are used only to visualize the zero-heavy UBE2H expression distribution.
pdac_cor <- cor.test(
  pdac_plot_df$aneuploidy_score_no_chr7q,
  pdac_plot_df$UBE2H_normalized,
  method = "spearman",
  exact = FALSE
)

pdac_cor

pdac_cor_label <- sprintf(
  "\u03c1 = %.2f\np %s",
  unname(pdac_cor$estimate),
  ifelse(
    pdac_cor$p.value < 2.2e-16,
    "< 2.2e-16",
    sprintf("= %.3g", pdac_cor$p.value)
  )
)

# Using fixed score groups only for visualization.
# Identical integer scores remain in the same group.
pdac_plot_df <- pdac_plot_df %>%
  mutate(
    score_group = case_when(
      aneuploidy_score_no_chr7q %in% 0:1 ~ "0–1",
      aneuploidy_score_no_chr7q == 2 ~ "2",
      aneuploidy_score_no_chr7q >= 3 ~ "3+",
      TRUE ~ NA_character_
    ),
    score_group = factor(
      score_group,
      levels = c("0–1", "2", "3+")
    )
  )

pdac_group_counts <- pdac_plot_df %>%
  count(
    score_group,
    name = "n_cells",
    .drop = FALSE
  )

# Confirming that all plotting cells were assigned to a group.
if (anyNA(pdac_plot_df$score_group)) {
  warning(
    sum(is.na(pdac_plot_df$score_group)),
    " cells were not assigned to a score group."
  )
}

pdac_plot_df <- pdac_plot_df %>%
  filter(!is.na(score_group))

# Checking number of cells contributing to each box.
pdac_group_counts %>%
  gt() %>%
  tab_header(
    title = "PDAC-A cells per aneuploidy-score group"
  ) %>%
  cols_label(
    score_group = "Aneuploidy-score group",
    n_cells = "Number of cells"
  )

# Boxplot.
p_pdac_score_group <- ggplot(
  pdac_plot_df,
  aes(
    x = score_group,
    y = UBE2H_normalized
  )
  ) +
  geom_boxplot(
    fill = "#B44FA3",
    color = "black",
    outlier.shape = NA,
    width = 0.6,
    linewidth = 0.6,
    alpha = 0.8
  ) +
  annotate(
    "text",
    x = -Inf,
    y = Inf,
    label = pdac_cor_label,
    hjust = -0.15,
    vjust = 1.5,
    size = 5
  ) +
  labs(
    x = "Aneuploidy score excluding 7q",
    y = "UBE2H Seurat-normalized expression",
    title = paste0(
      patient,
      ": UBE2H by aneuploidy score group"
    ),
    subtitle = "Cancer cells, chromosome 7q-neutral only"
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.text = element_text(
      color = "black",
      face = "bold"
    ),
    axis.title = element_text(face = "bold"),
    plot.title = element_text(
      face = "bold",
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      hjust = 0.5,
      size = 11,
      color = "grey30"
    )
  )

print(p_pdac_score_group)

ggsave(
  filename = here(
    "figures",
    paste0(
      output_prefix,
      "_UBE2H_by_aneuploidy_score_groups_cancer_7qneutral.svg"
    )
  ),
  plot = p_pdac_score_group,
  width = 7,
  height = 5.5
)

ggsave(
  filename = here(
    "figures",
    paste0(
      output_prefix,
      "_UBE2H_by_aneuploidy_score_groups_cancer_7qneutral.png"
    )
  ),
  plot = p_pdac_score_group,
  width = 7,
  height = 5.5,
  dpi = 600
)
```


```{r session-info}
sessionInfo()
```
