---
title: "GSE84465 CopyKAT and UBE2H analysis"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  run_copykat: false
---

```{r setup, message=FALSE, warning=FALSE}
library(here)
library(data.table)
library(Matrix)
library(tidyverse)
library(copykat)
library(Seurat)
library(ComplexHeatmap)
library(circlize)
library(grid)
library(ggpubr)
library(rstatix)
library(gt)
library(broom)
library(svglite)

patients <- c("BT_S1", "BT_S2", "BT_S4", "BT_S6")
exclude_arms <- c("13p", "14p", "15p", "21p", "22p")
fraction_threshold <- 0.80

# Edit if the repository layout changes.
copykat_rds <- here(
  "data", "GSE84465", "GSE84465_all_CopyKAT_results.rds"
)

counts_file_candidates <- c(
  here("data", "GSE84465", "GSE84465_GBM_All_data.csv.gz"),
  here("data", "GSE84465", "GSE84465_GBM_All_data.csv")
)

series_file <- here(
  "data", "GSE84465", "GSE84465_series_matrix.txt"
)

arm_coords_file <- here("data", "arm_coords.csv")

output_dir <- here("results", "GSE84465")
figure_dir <- here("figures")

dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

counts_file <- counts_file_candidates[file.exists(counts_file_candidates)][1]

if (is.na(counts_file)) {
  stop(
    "Count matrix not found. Expected one of:\n",
    paste(counts_file_candidates, collapse = "\n")
  )
}

copykat_cores <- 4L
```

# Read the raw count matrix

```{r read-counts}
message("Reading count matrix: ", counts_file)

counts_df <- read.table(
  counts_file,
  header = TRUE,
  row.names = 1,
  check.names = FALSE,
  sep = "",
  quote = "\"",
  comment.char = "",
  stringsAsFactors = FALSE
)

counts <- as.matrix(counts_df)
storage.mode(counts) <- "numeric"
rm(counts_df)
invisible(gc())

counts <- counts[!duplicated(rownames(counts)), , drop = FALSE]
counts <- counts[rowSums(counts) > 0, , drop = FALSE]

stopifnot(
  nrow(counts) > 1000,
  ncol(counts) > 1,
  all(is.finite(counts)),
  all(counts >= 0),
  all(counts == floor(counts))
)

cat(sprintf(
  "Raw count matrix: %d genes x %d cells\n",
  nrow(counts),
  ncol(counts)
))
```

# Parse GEO sample metadata

```{r parse-series-matrix}
if (!file.exists(series_file)) {
  stop("Series matrix not found: ", series_file)
}

series_lines <- readLines(series_file, warn = FALSE)

parse_geo_row <- function(line) {
  fields <- strsplit(line, "\t", fixed = TRUE)[[1]]
  gsub('^"|"$', "", fields)
}

sample_rows <- lapply(
  series_lines[grepl("^!Sample_", series_lines)],
  parse_geo_row
)

get_characteristic <- function(label) {
  matching_rows <- Filter(
    function(x) {
      length(x) > 1 &&
        x[1] == "!Sample_characteristics_ch1" &&
        startsWith(x[2], paste0(label, ":"))
    },
    sample_rows
  )

  if (length(matching_rows) != 1) {
    stop(
      "Expected exactly one metadata row for '",
      label,
      "' but found ",
      length(matching_rows),
      "."
    )
  }

  sub(
    paste0("^", label, ":\\s*"),
    "",
    matching_rows[[1]][-1]
  )
}

description_rows <- Filter(
  function(x) x[1] == "!Sample_description",
  sample_rows
)

cell_description_rows <- description_rows[
  vapply(
    description_rows,
    function(x) any(grepl("^[0-9]+\\.[A-H][0-9]+$", x[-1])),
    logical(1)
  )
]

if (length(cell_description_rows) != 1) {
  stop("Could not uniquely identify the GEO row containing cell IDs.")
}

cell_metadata <- data.frame(
  cell = cell_description_rows[[1]][-1],
  patient = get_characteristic("patient id"),
  plate = get_characteristic("plate id"),
  well = get_characteristic("well"),
  cell_type = get_characteristic("cell type"),
  neoplastic = get_characteristic("neoplastic"),
  tissue = get_characteristic("tissue"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

rownames(cell_metadata) <- cell_metadata$cell

missing_metadata <- setdiff(colnames(counts), rownames(cell_metadata))

if (length(missing_metadata) > 0) {
  stop(
    length(missing_metadata),
    " count-matrix cells are absent from the GEO metadata."
  )
}

cell_metadata <- cell_metadata[colnames(counts), , drop = FALSE]

stopifnot(
  identical(rownames(cell_metadata), colnames(counts))
)

cell_metadata <- cell_metadata %>%
  mutate(
    patient = factor(patient, levels = patients),
    input_group = case_when(
      tolower(neoplastic) == "regular" ~ "reference",
      tolower(neoplastic) == "neoplastic" ~ "observation",
      TRUE ~ NA_character_
    )
  )

cat("Cells per patient:\n")
print(table(cell_metadata$patient, useNA = "ifany"))

cat("\nNeoplastic annotation by patient:\n")
print(table(
  cell_metadata$patient,
  cell_metadata$neoplastic,
  useNA = "ifany"
))

stopifnot(
  all(patients %in% unique(as.character(cell_metadata$patient)))
)

group_counts <- cell_metadata %>%
  count(patient, input_group) %>%
  arrange(patient, input_group)

group_counts %>% gt()

for (p in patients) {
  n_ref <- sum(
    cell_metadata$patient == p &
      cell_metadata$input_group == "reference",
    na.rm = TRUE
  )

  n_obs <- sum(
    cell_metadata$patient == p &
      cell_metadata$input_group == "observation",
    na.rm = TRUE
  )

  if (n_ref < 5) {
    stop(p, " has fewer than five reference cells.")
  }

  if (n_obs < 1) {
    stop(p, " has no neoplastic observation cells.")
  }
}
```

# Run or load CopyKAT for each patient

Set `params$run_copykat: true` to run CopyKAT. Otherwise, the notebook loads the RDS file.

```{r copykat-four-patients}
copykat_results <- NULL

if (isTRUE(params$run_copykat)) {

  copykat_results <- setNames(
    vector("list", length(patients)),
    patients
  )

  for (p in patients) {
    message("Running CopyKAT for ", p)

    patient_cells <- rownames(cell_metadata)[
      cell_metadata$patient == p
    ]

    patient_counts <- counts[
      ,
      patient_cells,
      drop = FALSE
    ]

    patient_meta <- cell_metadata[
      patient_cells,
      ,
      drop = FALSE
    ]

    # Remove genes not detected in this patient.
    patient_counts <- patient_counts[
      rowSums(patient_counts) > 0,
      ,
      drop = FALSE
    ]

    # Light QC to avoid near-empty cells.
    detected_genes <- colSums(patient_counts > 0)
    library_size <- colSums(patient_counts)

    keep_cells <- (
      detected_genes >= 200 &
      library_size >= 500
    )

    patient_counts <- patient_counts[
      ,
      keep_cells,
      drop = FALSE
    ]

    patient_meta <- patient_meta[
      colnames(patient_counts),
      ,
      drop = FALSE
    ]

    norm_cells <- rownames(patient_meta)[
      patient_meta$input_group == "reference"
    ]

    cat(sprintf(
      "%s: %d genes x %d cells; %d regular reference cells\n",
      p,
      nrow(patient_counts),
      ncol(patient_counts),
      length(norm_cells)
    ))

    if (length(norm_cells) < 5) {
      stop(
        p,
        " has fewer than five regular reference cells after QC."
      )
    }

    patient_dir <- file.path(output_dir, p)

    dir.create(
      patient_dir,
      recursive = TRUE,
      showWarnings = FALSE
    )

    old_wd <- getwd()

    result <- tryCatch(
      {
        setwd(patient_dir)

        copykat::copykat(
          rawmat = patient_counts,
          id.type = "S",
          ngene.chr = 5,
          win.size = 25,
          KS.cut = 0.1,
          sam.name = p,
          distance = "euclidean",
          norm.cell.names = norm_cells,
          output.seg = "FALSE",
          plot.genes = "TRUE",
          n.cores = copykat_cores
        )
      },
      finally = {
        setwd(old_wd)
      }
    )

    copykat_results[[p]] <- result
  }

  # Save one combined four-patient object.
  dir.create(
    dirname(copykat_rds),
    recursive = TRUE,
    showWarnings = FALSE
  )

  saveRDS(
    copykat_results,
    copykat_rds
  )

} else {

  if (!file.exists(copykat_rds)) {
    stop(
      "Combined CopyKAT RDS not found: ",
      copykat_rds,
      ". Set params$run_copykat to true to generate it."
    )
  }

  copykat_results <- readRDS(
    copykat_rds
  )
}

missing_patients <- setdiff(
  patients,
  names(copykat_results)
)

if (length(missing_patients) > 0) {
  stop(
    "Combined CopyKAT object is missing: ",
    paste(missing_patients, collapse = ", ")
  )
}

for (p in patients) {
  required_components <- c("prediction", "CNAmat")

  missing_components <- setdiff(
    required_components,
    names(copykat_results[[p]])
  )

  if (length(missing_components) > 0) {
    stop(
      "CopyKAT result for ",
      p,
      " is missing: ",
      paste(missing_components, collapse = ", ")
    )
  }
}

cat("Loaded CopyKAT results for:\n")
print(names(copykat_results))
```


# Combine CopyKAT predictions

```{r combine-predictions}
pred_all <- bind_rows(lapply(patients, function(p) {
  pred <- as.data.frame(copykat_results[[p]]$prediction)

  pred %>%
    mutate(
      patient = p,
      cell_type = cell_metadata[cell.names, "cell_type"],
      neoplastic = cell_metadata[cell.names, "neoplastic"],
      input_group = cell_metadata[cell.names, "input_group"]
    )
}))

stopifnot(
  !anyNA(pred_all$patient),
  !anyNA(pred_all$cell.names)
)

cat("Overall CopyKAT calls:\n")
print(table(pred_all$copykat.pred, useNA = "ifany"))

cat("\nCopyKAT calls by patient:\n")
print(table(
  pred_all$patient,
  pred_all$copykat.pred,
  useNA = "ifany"
))

cat("\nCopyKAT calls by patient and input group:\n")
print(xtabs(
  ~ patient + input_group + copykat.pred,
  data = pred_all
))

write.table(
  pred_all,
  file.path(output_dir, "GSE84465_all_CopyKAT_predictions.txt"),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)

pred_all %>%
  count(patient, input_group, copykat.pred) %>%
  arrange(patient, input_group, copykat.pred) %>%
  gt(groupname_col = "patient") %>%
  tab_header(title = "CopyKAT predictions across four patients") %>%
  cols_label(
    input_group = "Input group",
    copykat.pred = "CopyKAT prediction",
    n = "Cell count"
  )
```

# Normalize UBE2H expression

```{r normalize-ube2h}
seu <- CreateSeuratObject(
  counts = as(counts, "dgCMatrix"),
  meta.data = cell_metadata,
  project = "GSE84465_GBM",
  min.cells = 0,
  min.features = 0
)

seu <- NormalizeData(
  seu,
  normalization.method = "LogNormalize",
  scale.factor = 10000,
  verbose = FALSE
)

ube2h_matches <- which(toupper(rownames(seu)) == "UBE2H")
stopifnot(length(ube2h_matches) == 1)

ube2h_id <- rownames(seu)[ube2h_matches]

ube2h_df <- FetchData(
  seu,
  vars = ube2h_id,
  layer = "data"
)

colnames(ube2h_df) <- "UBE2H_normalized"
ube2h_df$cell <- rownames(ube2h_df)
```

# Read chromosome-arm coordinates

```{r arm-coordinates}
if (!file.exists(arm_coords_file)) {
  stop("Chromosome-arm coordinate file not found: ", arm_coords_file)
}

arm_coords <- read_csv(
  arm_coords_file,
  show_col_types = FALSE
)

required_arm_cols <- c("chrom_num", "arm", "start", "end")

stopifnot(
  all(required_arm_cols %in% colnames(arm_coords))
)
```

# Extract CN values and derive patient-specific thresholds

```{r patient-thresholds}
cn_values_list <- setNames(vector("list", length(patients)), patients)
cn_annotation_list <- setNames(vector("list", length(patients)), patients)

threshold_rows <- vector("list", length(patients))
signal_rows <- vector("list", length(patients))

for (i in seq_along(patients)) {
  p <- patients[[i]]

  cn <- as.data.frame(
    copykat_results[[p]]$CNAmat,
    check.names = FALSE
  )

  # Reverse syntactic-name conversion applied by R to numeric-leading IDs.
  colnames(cn) <- sub(
    "^X(?=[0-9])",
    "",
    colnames(cn),
    perl = TRUE
  )

  required_annotation_cols <- c("chrom", "chrompos", "abspos")

  stopifnot(
    all(required_annotation_cols %in% colnames(cn))
  )

  patient_pred <- pred_all %>%
    filter(patient == p)

  copykat_cell_cols <- intersect(
    patient_pred$cell.names,
    colnames(cn)
  )

  if (length(copykat_cell_cols) == 0) {
    stop("No CopyKAT CN columns matched predictions for ", p, ".")
  }

  cn_values <- as.matrix(
    cn[, copykat_cell_cols, drop = FALSE]
  )

  storage.mode(cn_values) <- "numeric"

  diploid_cells <- patient_pred$cell.names[
    patient_pred$copykat.pred == "diploid"
  ]

  diploid_cells <- intersect(
    diploid_cells,
    colnames(cn_values)
  )

  if (length(diploid_cells) == 0) {
    stop("No CopyKAT-predicted diploid cells found for ", p, ".")
  }

  diploid_signal <- as.numeric(
    cn_values[, diploid_cells, drop = FALSE]
  )

  diploid_signal <- diploid_signal[is.finite(diploid_signal)]

  bin_signal_threshold <- unname(
    quantile(
      abs(diploid_signal),
      probs = 0.90,
      na.rm = TRUE
    )
  )

  cn_values_list[[p]] <- cn_values
  cn_annotation_list[[p]] <- cn

  threshold_rows[[i]] <- tibble(
    patient = p,
    n_diploid_cells = length(diploid_cells),
    bin_signal_threshold = bin_signal_threshold
  )

  signal_rows[[i]] <- tibble(
    patient = p,
    signal = diploid_signal,
    bin_signal_threshold = bin_signal_threshold
  )

  cat(sprintf(
    "%s: %d bins x %d cells; threshold = %.6f\n",
    p,
    nrow(cn_values),
    ncol(cn_values),
    bin_signal_threshold
  ))
}

threshold_df <- bind_rows(threshold_rows) %>%
  mutate(patient = factor(patient, levels = patients))

signal_df <- bind_rows(signal_rows) %>%
  mutate(patient = factor(patient, levels = patients))

threshold_df %>% gt()
```

```{r threshold-distributions, fig.width=10, fig.height=7}
ggplot(signal_df, aes(x = signal)) +
  geom_histogram(
    bins = 150,
    boundary = 0
  ) +
  geom_vline(
    data = threshold_df,
    aes(xintercept = bin_signal_threshold),
    linetype = "dashed",
    linewidth = 0.8
  ) +
  geom_vline(
    data = threshold_df,
    aes(xintercept = -bin_signal_threshold),
    linetype = "dashed",
    linewidth = 0.8
  ) +
  facet_wrap(
    ~ patient,
    scales = "free_y",
    ncol = 2
  ) +
  labs(
    title = "Distribution of CopyKAT bin signals in diploid cells",
    subtitle = paste0(
      "Dashed lines indicate each patient's 90th percentile ",
      "of absolute diploid-cell signal"
    ),
    x = "CopyKAT bin signal",
    y = "Bin-cell observations"
  ) +
  theme_classic()
```

# Convert CopyKAT bins to chromosome-arm calls

An arm is called gained when at least 80% of its bins exceed the positive 
patient-specific threshold and lost when at least 80% fall below the negative 
threshold.

```{r arm-calls-four-patients}
bin_signal_thresholds <- threshold_df %>%
  mutate(patient = as.character(patient)) %>%
  select(patient, bin_signal_threshold) %>%
  tibble::deframe()

missing_thresholds <- setdiff(
  patients,
  names(bin_signal_thresholds)
)

if (length(missing_thresholds) > 0) {
  stop(
    "Missing thresholds for: ",
    paste(missing_thresholds, collapse = ", ")
  )
}

arm_call_list <- setNames(vector("list", length(patients)), patients)
bin_arm_list <- setNames(vector("list", length(patients)), patients)
arm_call_qc <- setNames(vector("list", length(patients)), patients)

for (p in patients) {
  cn <- cn_annotation_list[[p]]
  cn_values <- cn_values_list[[p]]
  bin_signal_threshold <- unname(bin_signal_thresholds[[p]])

  bin_annot <- cn %>%
    transmute(
      bin_id = seq_len(nrow(cn)),
      chrom = suppressWarnings(as.numeric(as.character(chrom))),
      chrompos = as.numeric(chrompos),
      abspos = as.numeric(abspos)
    ) %>%
    filter(chrom %in% 1:22)

  bin_arm <- bin_annot %>%
    inner_join(
      arm_coords %>%
        transmute(
          chrom_num = as.numeric(chrom_num),
          arm,
          arm_start = start,
          arm_end = end
        ),
      by = join_by(
        chrom == chrom_num,
        chrompos >= arm_start,
        chrompos < arm_end
      )
    ) %>%
    filter(!arm %in% exclude_arms) %>%
    select(
      bin_id,
      chrom,
      chrompos,
      abspos,
      arm
    ) %>%
    arrange(chrom, chrompos)

  if (nrow(bin_arm) == 0) {
    stop("No bins were assigned to chromosome arms for ", p, ".")
  }

  assignment_qc <- bin_arm %>%
    count(bin_id, name = "n_assignments") %>%
    summarise(
      max_assignments = max(n_assignments),
      duplicated_bins = sum(n_assignments > 1)
    )

  stopifnot(
    assignment_qc$max_assignments <= 1
  )

  arms <- unique(bin_arm$arm)

  arm_call <- matrix(
    0L,
    nrow = length(arms),
    ncol = ncol(cn_values),
    dimnames = list(arms, colnames(cn_values))
  )

  for (a in arms) {
    bin_ids <- bin_arm$bin_id[bin_arm$arm == a]

    x <- cn_values[
      bin_ids,
      ,
      drop = FALSE
    ]

    frac_gain <- colMeans(
      x > bin_signal_threshold,
      na.rm = TRUE
    )

    frac_loss <- colMeans(
      x < -bin_signal_threshold,
      na.rm = TRUE
    )

    arm_call[
      a,
      frac_gain >= fraction_threshold
    ] <- 1L

    arm_call[
      a,
      frac_loss >= fraction_threshold
    ] <- -1L
  }

  stopifnot(
    !any(rownames(arm_call) %in% exclude_arms),
    all(arm_call %in% c(-1L, 0L, 1L))
  )

  arm_call_list[[p]] <- arm_call
  bin_arm_list[[p]] <- bin_arm

  arm_call_qc[[p]] <- tibble(
    patient = p,
    bin_signal_threshold = bin_signal_threshold,
    fraction_threshold = fraction_threshold,
    total_copykat_bins = nrow(cn_values),
    assigned_bins = nrow(bin_arm),
    arms = nrow(arm_call),
    cells = ncol(arm_call),
    gain_calls = sum(arm_call == 1L),
    loss_calls = sum(arm_call == -1L),
    neutral_calls = sum(arm_call == 0L)
  )
}

arm_call_qc_df <- bind_rows(arm_call_qc)
arm_call_qc_df %>% gt()
```

# Build combined per-cell metadata and scores

The variable `aneuploidy_score_no_7q` excludes 7q only

```{r combined-final-meta}
score_df <- bind_rows(lapply(patients, function(p) {
  arm_call <- arm_call_list[[p]]

  tibble(
    cell = colnames(arm_call),
    patient = p,
    aneuploidy_score = colSums(abs(arm_call)),
    aneuploidy_score_no_7q = colSums(abs(
      arm_call[
        !rownames(arm_call) %in% "7q",
        ,
        drop = FALSE
      ]
    )),
    chr7q_call = if ("7q" %in% rownames(arm_call)) {
      as.integer(arm_call["7q", ])
    } else {
      NA_integer_
    }
  )
})) %>%
  mutate(
    patient = factor(patient, levels = patients),
    chr7q_status = factor(
      chr7q_call,
      levels = c(-1, 0, 1),
      labels = c("Loss", "Neutral", "Gain")
    )
  )

final_meta <- pred_all %>%
  transmute(
    cell = cell.names,
    patient,
    cell_type,
    neoplastic,
    input_group,
    copykat_call = copykat.pred
  ) %>%
  left_join(ube2h_df, by = "cell") %>%
  left_join(score_df, by = c("cell", "patient"))

stopifnot(
  !anyNA(final_meta$cell),
  !anyDuplicated(final_meta$cell)
)

write.table(
  final_meta,
  file.path(output_dir, "GSE84465_CopyKAT_per_cell_metadata.txt"),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)

final_meta %>%
  group_by(patient, input_group, copykat_call) %>%
  summarise(
    n = n(),
    median_score = median(aneuploidy_score, na.rm = TRUE),
    mean_score = mean(aneuploidy_score, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  gt()
```


# Build the core 7q-neutral analysis dataset

This shared table contains neoplastic, chromosome 7q-neutral cells with
non-missing UBE2H expression and aneuploidy scores. It is used by the pooled
boxplot and patient-summary line plot.

```{r core-7q-neutral-data}
core_7q_neutral_df <- final_meta %>%
  filter(
    input_group == "observation",
    chr7q_status == "Neutral",
    !is.na(aneuploidy_score_no_7q),
    !is.na(UBE2H_normalized),
    !is.na(patient)
  ) %>%
  mutate(patient = factor(patient, levels = patients))

core_7q_neutral_df %>%
  count(patient, name = "n_cells") %>%
  gt()
```

# Pooled Spearman correlation and fixed-score-group boxplot

```{r pooled-binned-boxplot, fig.width=6.5, fig.height=6}
pooled_df <- core_7q_neutral_df

# Naive pooled Spearman correlation using the original continuous score.
pooled_cor_test <- cor.test(
  pooled_df$aneuploidy_score_no_7q,
  pooled_df$UBE2H_normalized,
  method = "spearman",
  exact = FALSE
)

pooled_rho <- unname(pooled_cor_test$estimate)
pooled_p <- pooled_cor_test$p.value
pooled_n_cells <- nrow(pooled_df)
pooled_n_patients <- n_distinct(pooled_df$patient)

pooled_label <- paste0(
  "\u03c1 = ", sprintf("%.2f", pooled_rho),
  "\np = ",
  format.pval(
    pooled_p,
    digits = 3,
    eps = 2.2e-16
  ),
  "\n", pooled_n_cells,
  " cells from ",
  pooled_n_patients,
  " patients"
)

pooled_score_group_df <- pooled_df %>%
  mutate(
    score_group = case_when(
      aneuploidy_score_no_7q %in% 0:1 ~ "0–1",
      aneuploidy_score_no_7q %in% 2:3 ~ "2–3",
      aneuploidy_score_no_7q %in% 4:5 ~ "4–5",
      aneuploidy_score_no_7q >= 6 ~ "6+",
      TRUE ~ NA_character_
    ),
    score_group = factor(
      score_group,
      levels = c("0–1", "2–3", "4–5", "6+")
    )
  ) %>%
  filter(!is.na(score_group))

# Check pooled and patient-specific composition of each score group.
pooled_score_group_counts <- pooled_score_group_df %>%
  count(
    score_group,
    name = "n_cells",
    .drop = FALSE
  )

pooled_score_group_counts %>% gt()

pooled_score_group_patient_counts <- pooled_score_group_df %>%
  count(
    patient,
    score_group,
    name = "n_cells",
    .drop = FALSE
  )

pooled_score_group_patient_counts %>% gt()

p_pooled_score_group <- ggplot(
  pooled_score_group_df,
  aes(
    x = score_group,
    y = UBE2H_normalized
  )
) +
  geom_boxplot(
    fill = "#B44FA3",
    color = "black",
    outlier.shape = NA,
    width = 0.62,
    linewidth = 0.6
  ) +
  annotate(
    "text",
    x = -Inf,
    y = Inf,
    label = pooled_label,
    hjust = -0.08,
    vjust = 1.15,
    size = 4.5
  ) +
  labs(
    x = "Aneuploidy score excluding 7q",
    y = "UBE2H Seurat-normalized expression",
    title = "UBE2H by pooled aneuploidy score group",
    subtitle = paste0(
      "Neoplastic, chromosome 7q-neutral cells ",
      "from four patients"
    )
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.text = element_text(
      color = "black",
      face = "bold"
    ),
    axis.title = element_text(face = "bold"),
    plot.title = element_text(
      face = "bold",
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      hjust = 0.5,
      size = 10.5
    )
  )

p_pooled_score_group

ggsave(
  filename = file.path(
    figure_dir,
    "GSE84465_pooled_UBE2H_by_aneuploidy_score_group_7qneutral.svg"
  ),
  plot = p_pooled_score_group,
  width = 6.5,
  height = 6
)
```

```{r combined-patient-summary-line-plot, fig.width=8, fig.height=6}
combined_patient_df <- core_7q_neutral_df %>%
  mutate(
    score_group = case_when(
      aneuploidy_score_no_7q %in% 0:1 ~ "0–1",
      aneuploidy_score_no_7q %in% 2:3 ~ "2–3",
      aneuploidy_score_no_7q %in% 4:5 ~ "4–5",
      aneuploidy_score_no_7q >= 6 ~ "6+",
      TRUE ~ NA_character_
    ),
    score_group = factor(
      score_group,
      levels = c("0–1", "2–3", "4–5", "6+")
    ),
    patient = factor(
      patient,
      levels = patients
    )
  ) %>%
  filter(!is.na(score_group))

# Calculate patient-specific mean expression and SEM
patient_summary_df <- combined_patient_df %>%
  group_by(patient, score_group) %>%
  summarise(
    n_cells = n(),
    mean_expression = mean(
      UBE2H_normalized,
      na.rm = TRUE
    ),
    sd_expression = sd(
      UBE2H_normalized,
      na.rm = TRUE
    ),
    sem_expression = sd_expression / sqrt(n_cells),
    .groups = "drop"
  ) %>%
  mutate(
    lower_sem = mean_expression - sem_expression,
    upper_sem = mean_expression + sem_expression
  )

patient_summary_df %>% gt()

# Four distinct shades of grey
patient_greys <- c(
  "BT_S1" = "#222222",
  "BT_S2" = "#555555",
  "BT_S4" = "#888888",
  "BT_S6" = "#C4C4C4"
)

p_combined_patient_mean_sem <- ggplot(
  patient_summary_df,
  aes(
    x = score_group,
    y = mean_expression,
    group = patient,
    color = patient,
    shape = patient
  )
) +
  geom_line(
    linewidth = 0.9,
    alpha = 0.9
  ) +
  geom_errorbar(
    aes(
      ymin = lower_sem,
      ymax = upper_sem
    ),
    width = 0.10,
    linewidth = 0.65,
    alpha = 0.45
  ) +
  geom_point(
    size = 3,
    stroke = 0.8
  ) +
  scale_color_manual(
    values = patient_greys
  ) +
  labs(
    x = "Aneuploidy score excluding 7q",
    y = "Mean UBE2H Seurat-normalized expression",
    title = "UBE2H expression by aneuploidy score group",
    subtitle = paste0(
      "Patient-specific means ± SEM; ",
      "neoplastic, chromosome 7q-neutral cells"
    ),
    color = "Patient",
    shape = "Patient"
  ) +
  theme_classic(base_size = 13) +
  theme(
    axis.text = element_text(color = "black"),
    axis.title = element_text(face = "bold"),
    plot.title = element_text(
      face = "bold",
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      hjust = 0.5,
      size = 10.5
    ),
    legend.position = "right"
  )

p_combined_patient_mean_sem

ggsave(
  filename = file.path(
    figure_dir,
    "GSE84465_UBE2H_by_aneuploidy_score_group_patient_mean_SEM_7qneutral.svg"
  ),
  plot = p_combined_patient_mean_sem,
  width = 8,
  height = 6
)
```


# Patient-adjusted regression

```{r optional-patient-adjusted-regression}
regression_df <- core_7q_neutral_df %>%
  mutate(patient = factor(patient, levels = patients))

fit_patient_adjusted <- lm(
  UBE2H_normalized ~ aneuploidy_score_no_7q + patient,
  data = regression_df
)

summary(fit_patient_adjusted)

broom::tidy(
  fit_patient_adjusted,
  conf.int = TRUE
) %>%
  gt()
```


# Diagnostic: bin-level CopyKAT heatmaps

```{r bin-heatmaps, fig.width=13, fig.height=8}
copykat_col_fun <- circlize::colorRamp2(
  c(-1, 0, 1),
  c("#3B6FB6", "white", "#C74440")
)

group_colors <- c(
  "Regular reference" = "#7A7A7A",
  "Neoplastic: diploid" = "#69A45A",
  "Neoplastic: aneuploid" = "#D18B2C",
  "Not defined" = "#BDBDBD"
)

for (p in patients) {
  cn_df <- cn_annotation_list[[p]]

  patient_cells <- pred_all$cell.names[
    pred_all$patient == p
  ]

  cell_cols <- intersect(
    colnames(cn_df),
    patient_cells
  )

  cn_mat_plot <- as.matrix(
    cn_df[, cell_cols, drop = FALSE]
  )

  storage.mode(cn_mat_plot) <- "numeric"

  bin_chr <- as.character(cn_df$chrom)
  keep_bins <- suppressWarnings(as.numeric(bin_chr)) %in% 1:22

  cn_mat_plot <- cn_mat_plot[keep_bins, , drop = FALSE]
  bin_chr <- bin_chr[keep_bins]

  cell_meta <- final_meta %>%
    filter(
      patient == p,
      cell %in% colnames(cn_mat_plot)
    ) %>%
    select(
      cell,
      cell_type,
      input_group,
      copykat_call,
      aneuploidy_score
    ) %>%
    mutate(
      group = case_when(
        input_group == "reference" ~ "Regular reference",
        input_group == "observation" &
          copykat_call == "diploid" ~ "Neoplastic: diploid",
        input_group == "observation" &
          copykat_call == "aneuploid" ~ "Neoplastic: aneuploid",
        TRUE ~ "Not defined"
      ),
      group = factor(group, levels = names(group_colors))
    )

  rownames(cell_meta) <- cell_meta$cell
  cell_meta <- cell_meta[colnames(cn_mat_plot), , drop = FALSE]

  cell_order <- cell_meta %>%
    arrange(group, aneuploidy_score) %>%
    pull(cell)

  cn_mat_plot <- cn_mat_plot[, cell_order, drop = FALSE]
  cell_meta <- cell_meta[cell_order, , drop = FALSE]

  bin_chr_factor <- factor(
    bin_chr,
    levels = as.character(1:22)
  )

  stopifnot(
    ncol(cn_mat_plot) > 0,
    nrow(cn_mat_plot) > 0,
    identical(rownames(cell_meta), colnames(cn_mat_plot)),
    length(bin_chr_factor) == nrow(cn_mat_plot),
    !anyNA(cell_meta$group),
    !anyNA(bin_chr_factor)
  )

  top_anno <- HeatmapAnnotation(
    Group = cell_meta$group,
    col = list(Group = group_colors),
    annotation_name_side = "left"
  )

  ht <- Heatmap(
    cn_mat_plot,
    name = "CopyKAT",
    col = copykat_col_fun,
    top_annotation = top_anno,
    cluster_rows = FALSE,
    row_split = bin_chr_factor,
    cluster_row_slices = FALSE,
    row_gap = unit(0.4, "mm"),
    cluster_columns = FALSE,
    show_row_names = FALSE,
    show_column_names = FALSE,
    row_title = as.character(1:22),
    row_title_rot = 0,
    row_title_gp = gpar(fontsize = 8),
    column_split = cell_meta$group,
    column_gap = unit(2, "mm"),
    rect_gp = gpar(col = NA),
    column_title = paste0(p, " cells"),
    use_raster = FALSE
  )

  ComplexHeatmap::draw(
    ht,
    heatmap_legend_side = "right",
    annotation_legend_side = "right"
  )
}
```


# Diagnostic: arm-call heatmaps

```{r arm-heatmaps, fig.width=13, fig.height=8}
arm_order <- as.vector(rbind(
  paste0(1:22, "p"),
  paste0(1:22, "q")
))

call_colors <- c(
  `-1` = "#3B6FB6",
  `0` = "white",
  `1` = "#C74440"
)

for (p in patients) {
  arm_call <- arm_call_list[[p]]

  current_order <- arm_order[
    arm_order %in% rownames(arm_call)
  ]

  arm_call_plot <- arm_call[
    current_order,
    ,
    drop = FALSE
  ]

  heatmap_meta <- final_meta %>%
    filter(
      patient == p,
      cell %in% colnames(arm_call_plot)
    ) %>%
    select(
      cell,
      input_group,
      copykat_call,
      aneuploidy_score
    ) %>%
    mutate(
      group = case_when(
        input_group == "reference" ~ "Regular reference",
        input_group == "observation" &
          copykat_call == "diploid" ~ "Neoplastic: diploid",
        input_group == "observation" &
          copykat_call == "aneuploid" ~ "Neoplastic: aneuploid",
        TRUE ~ "Not defined"
      ),
      group = factor(group, levels = names(group_colors))
    )

  rownames(heatmap_meta) <- heatmap_meta$cell

  heatmap_meta <- heatmap_meta[
    colnames(arm_call_plot),
    ,
    drop = FALSE
  ]

  cell_order <- heatmap_meta %>%
    arrange(group, aneuploidy_score) %>%
    pull(cell)

  arm_call_plot <- arm_call_plot[, cell_order, drop = FALSE]
  heatmap_meta <- heatmap_meta[cell_order, , drop = FALSE]

  top_anno <- HeatmapAnnotation(
    Group = heatmap_meta$group,
    `Arm score` = anno_barplot(
      heatmap_meta$aneuploidy_score,
      border = FALSE,
      height = unit(1.5, "cm")
    ),
    col = list(Group = group_colors),
    annotation_name_side = "left"
  )

  ht <- Heatmap(
    arm_call_plot,
    name = "Arm call",
    col = call_colors,
    top_annotation = top_anno,
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    show_row_names = TRUE,
    show_column_names = FALSE,
    row_names_side = "left",
    row_names_gp = gpar(fontsize = 9),
    column_split = heatmap_meta$group,
    column_gap = unit(2, "mm"),
    rect_gp = gpar(col = NA),
    heatmap_legend_param = list(
      at = c(-1, 0, 1),
      labels = c("Loss", "Neutral", "Gain")
    ),
    column_title = paste0(p, " cells"),
    row_title = "Chromosome arm"
  )

  ComplexHeatmap::draw(
    ht,
    heatmap_legend_side = "right",
    annotation_legend_side = "right"
  )
}
```

# Session information

```{r session-info}
sessionInfo()
```
